\Huge \bf \sc Design and implementation of sensor fusion and control for an autonomous quadrotor

### 2013-11-21 latex/paper_largo/paper_largo.tex
papers actualizados

%\vspace*{-.5cm}
\begin{abstract}
%\boldmath
This paper describes the design and integration of an instrumentation and control system that allows the autonomous flight of a quadrotor. A comercial frame is used, a mathematical model for the quadrotor is developed and its parameters determined from the characterization of the unit. An intelligence is integrated to play the role of the flight controller. A 9 degrees of freedom Inertial Measurement Unit (IMU) equipped with a barometer is calibrated and added to the platform. Data from the IMU is combined with the information provided by a GPS within a modified Extended Kalman Filter (EKF) to obtain a reliable estimation of the state variables. The control actions are obtained from a proportional-integral controller based on an improved Linear Quadratic Regulator (LQR) algorithm. As shown in Section \refp{sec:results-tests}, a stable autonomous platform is achieved.%For windless environments, a stable platform is achieved.
% -cambio-
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

There is an important growth in the interest about unmanned aerial vehicles (UAV) in relation to its capabilities to perform a wide spectrum of tasks as monitoring, surveillance, aerial photography, exploration, delivery, rescue, remote sensing, etc.
The continuous advances in low power MEMS sensor devices, low power embedded processors, energy storage through efficient electric batteries and electrical motors technologies have been driving forces in the process of implementation of new and more sophisticated devices of this kind, particularly miniature flying robots. Quadrotors UAV is one of the most popular architectures.
They emerged also as a typical platform for research. There is much research work about path planning and vision towards navigation in unstructured scenarios (e.g.: \cite{bib:tio_cuatro, bib:tio_cinco, bib:tio_seis}), fault tolerant navigation (e.g.:\cite{bib:tio_tres}), path tracking (e.g.: \cite{bib:tio_ocho}), control techniques for UAV such as neural networks \cite{bib:tio_uno}, gain-schedulling \cite{bib:tio_tres}, feedback linearization (e.g.:\cite{bib:tio_seis, bib:tio_siete}), PI control \cite{bib:tio_ocho}, Adaptive Control \cite{bib:tio_dos}, LQR (e.g.: \cite{bib:tio_cinco, bib:lqrnotes}), backstepping control (e.g.:\cite{bib:tio_cuatro}, \cite{bib:tio_seis}), etc.
If state estimation, path planning or control techniques are to be experimented in a physical setup, an important issue is to have the possibility to gain access to the different blocks of the navigational system. 
The purpose of this work is to develop the instrumentation and control of an autonomous quadracopter in which is possible to modify and test different state estimation techniques, path planning algorithms and control laws. The solution was reached using a commercial mechanical platform, and design and build an ad-hoc instrumentation and control navigation system.
An aerial vehicle is inherently unstable, staying still is not a simple task. This paper focuses on the stabilization of a quadrotor, explaining the development of the mathematical model to represent the system, the filtering techniques applied for sensor data fusion and the control system that allows the quadrotor to fly.

\begin{figure}
%\vspace{-.8cm}
	\centering
	\includegraphics[width=.8\columnwidth]{./pics_paper/javi.png}
	\caption{\textbf{Commercial quadrotor used.}}
	\label{fig:commercial_quadrotor}
\end{figure}

\begin{figure}[h!]
\vspace{-.3cm}
	\centering
	\includegraphics[width=.8\columnwidth]{./pics_paper/diagrama_general.png}
	\caption{\textbf{General diagram.}}
	\label{fig:diagrama_tio}
\end{figure}

\section{System architecture}
The platform is based on the commercial radio controlled quadrotor shown in Figure \refp{fig:commercial_quadrotor}. The length between opposite propellers is $61.5cm$, the weight is $990g$, and it has $1300g$ of payload. The frame and the Electronic Speed Controllers (ESCs) used for the motors were preserved, whereas the IMU and intelligence were replaced by the flight controller that was developed. A BeagleBoard\footnote{BeagleBoard development board - \url{http://beagleboard.org/}} running Linux\footnote{Angstrom distribution: \url{http://www.angstrom-distribution.org/}} performs the computations required to convert raw data received from the IMU\footnote{\textit{Mongoose} IMU - \url{http://store.ckdevices.com/}} over a UART and combine it using an EKF. The EKF overcomes the problems inherent to each sensor and filters out noise, providing a reliable estimation of the state vector. Once the current state is known, the LQR algorithm is used to derive the control actions required to bring the system to the desired set-point. A complete diagram of the implemented system is shown in Figure \refp{fig:diagrama_tio}.

% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds

%\hfill January 11, 2007
The two main goals are to integrate additional sensors and intelligence to the available platform to obtain a state estimation, and design and integrate a control system that, using the state estimation, achieves the autonomous flight.

\section{Model of a quadrotor}
\label{sec:modelo}

\subsection{Definitions}
\label{sec:modelo-defs}

A diagram of the quadrotor is shown in Figure \refp{fig:quad}. Two of the motors rotate clockwise (2 and 4) and the other two (1 and 3) rotate counterclockwise. This configuration allows the quadrotor to rotate, tilt and gain/lose altitud by setting different speeds on each motor. Two frames of reference (Figure \refp{fig:quad}) are constantly used through out this paper: an intertial frame $S_I - \lbrace \hat{i},\hat{j},\hat{k}\rbrace$ ($\lbrace \vec{x},\vec{y},\vec{z}\rbrace$), relative to the Earth, mapped to  North, West and Up respectively, and a non-inertial frame $S_q - \lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$ ($\lbrace \vec{x}_q,\vec{y}_q,\vec{z}_q\rbrace$) relative to the quadrotor. The mapping of one frame to the other can be achieved by applying the three rotations shown in Figure \refp{fig:rotaciones}. The angles $\lbrace \theta, \varphi, \psi\rbrace$ are known as Euler angles.

\begin{figure}[h!]
	\centering
	\includegraphics[width=.7\columnwidth]{./pics_paper/quad_coord.jpg}
	\vspace{-5pt}
	\caption{\textbf{Model of the quadrotor -} The blue arrows represent the inertial reference frame $S_I$, and the red arrows represent the non-inertial reference frame $S_q$. The cyan ``looped'' arrows indicate the direction of rotation of each motor, which rotate at $\omega_i$ and generate a torque $M_i$ opposite to their direction of rotation. The arrows labeled $T_{[1,2,3,4]}$ represent the thrust of the motors. The semicircle and the two yellow spheres indicate the $x_q$ axis of the unit.}
    %\vspace{-10pt}
	\label{fig:quad}
\end{figure}

\subsection{Dynamics-kinematics of the system}
\label{sec:modelo-dyn-kin}

From a detailed analysis of the dynamics and kinematics of the quadrotor, the equations \refp{eq:modelo} are obtained, and the state vector shown in \refp{eq:sv} is built to describe the system at any given time. The variables with subscript \textit{q} are referenced to the quadrotor frame $S_q$, the rest are relative to $S_I$:
\begin{equation}
\small
\boxed{\begin{aligned}&\dot{x}\!=\!v_{q_x} \cos \varphi \cos \theta \!+\! v_{q_y} ( \cos \theta \sin \varphi \sin \psi\!-\!\cos \varphi \sin \theta )\\
&\quad \!+\! v_{q_z}(\sin \psi \sin \theta \!+\! \cos \psi \cos \theta \sin \varphi)\\
&\dot{y}\!=\!v_{q_x} \cos \varphi \sin \theta \!+\! v_{q_y} (\cos \psi \cos \theta \!+\! \sin \theta \sin \varphi \sin \psi)\\
&\quad\!+\! v_{q_z}( \cos \psi \sin \theta \sin \varphi\!-\!\cos \theta \sin \psi )\\
&\dot{z}\!=\! \!-\!v_{q_x} \sin \varphi  \!+\! v_{q_y} \cos \varphi \sin \psi  \!+\! v_{q_z}\cos \varphi \cos \psi\\
&\dot{\psi}\!=\!\omega_{q_x} \!+\! \omega_{q_z}\tan\varphi \cos\psi \!+\! \omega_{q_y}\tan\varphi \sin\psi\\
&\dot{\varphi}\!=\!\omega_{q_y}\cos \psi \!-\! \omega_{q_z}\sin\psi\\
&\dot{\theta}\!=\!\omega_{q_z} \frac{\cos\psi}{\cos\varphi}  \!+\! \omega_{q_y}\frac{\sin\psi}{\cos\varphi}\\
&\dot{v_{q_x}}\!=\!v_{q_y} \omega_{q_z} \!-\! v_{q_z} \omega_{q_y}\!+\!g\sin\varphi\\
&\dot{v_{q_y}}\!=\!v_{q_z} \omega_{q_x} \!-\! v_{q_x} \omega_{q_z}\!-\!g\cos\varphi\sin\psi\\
&\dot{v_{q_z}}\!=\!v_{q_x} \omega_{q_y} \!-\! v_{q_y} \omega_{q_x}\!-\!g\cos\varphi\cos\psi\!+\!\frac{1}{M}\sum_{i\!=\!1}^4T_i\\
&\dot{\omega_{q_x}}\!=\!\frac{1}{I_{xx}}\omega_{q_y}\omega_{q_z}(I_{yy}\!-\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{xx}}\omega_{q_y}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{xx}}dMg\cos\varphi\sin\psi\!+\!\frac{1}{I_{xx}}L(T_2\!-\!T_4)\\
&\dot{\omega_{q_y}}\!=\!\frac{1}{I_{yy}}\omega_{q_x}\omega_{q_z}(\!-\!I_{xx}\!+\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{yy}}\omega_{q_x}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{yy}}dMg\sin\varphi\!+\!\frac{1}{I_{yy}}L(T_3\!-\!T_1)\\
&\dot{\omega_{q_z}}\!=\!\frac{1}{I_{zz}}\left(\!-\!Q_1\!+\!Q_2\!-\!Q_3\!+\!Q_4\right)\\
%\frac{\!-\!I_{zz_m}(\dot{w_1}\!-\!\dot{w_2}\+!\!\dot{w_3}\!-\!\dot{w_4})}{I_{zz}}\!+\!
\end{aligned}}
\label{eq:modelo}
\end{equation}
\begin{equation}
\vec{X}=\left\lbrace  x,y,z, \theta,\varphi,\psi, v_{q_z},v_{q_y},v_{q_z},\omega_{q_x},\omega_{q_y},\omega_{q_z} \right\rbrace
\label{eq:sv}
\end{equation}
where:
\begin{itemize}
\item $\lbrace x,y,z \rbrace$ represent the position of the center of mass of the system in $S_I$.
\item $\lbrace\theta,\varphi,\psi\rbrace$ are the Euler angles shown in Figure \refp{fig:rotaciones}.
\item $\lbrace v_{q_x},v_{q_y},v_{q_z}\rbrace$ are the linear velocities relative to $S_q$.
\item $\lbrace w_{q_x},w_{q_y},w_{q_z}\rbrace$ are the angular velocities relative to $S_q$ (right hand rule applied on $\lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$).
\end{itemize}

\begin{figure}[h!]
  \centering
  \vspace{-20pt}
  \subfloat[Rotation 1: Axis $\hat{k}$]{\label{fig:angulos_1}
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_1.png}}
  \subfloat[Rotation 2: Axis $\hat{j}$]{\label{fig:angulos_2} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_2.png}}
  \subfloat[Rotation 3: Axis $\hat{i}$]{\label{fig:angulos_3} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_3.png}}
  \caption{\textbf{Mapping -} Rotations applied on $S_I$ to obtain $S_q$.}
  \label{fig:rotaciones}
\end{figure}

The mathematical model developed is similar to the ones presented in \cite{bib:quadrotor-bible, bib:curso_quad}, but also takes into account that the center of gravity of the quadrotor is not at the same height as the propellers. Thus a momentum produced by the gravity force has to be added. While in \cite{bib:curso_quad} the linear velocities are expressed in an inertial frame, in this work are referenced to the quadrotor frame. This choice simplifies the theoretical development and the interpretation of the data provided by the IMU, which is mounted on the quadrotor and hence provides provides accelerations and angular velocities that are relative to $S_q$.

@@

The IMU is equiped with the following sensors:
\begin{itemize}
\item \textbf{Barometer:} Measures the absolute pressure of the environment. Variations of pressure are used to estimate variations in the altitude of the system.
\item \textbf{Thermometer:} The barometer includes a thermometer. The temperature data is used to apply a temperature compensation to the calibrations performed on the gyroscope and the accelerometer.
\item \textbf{Gyroscope:} A 3-axis gyroscope is used to measure angular velocity of $S_q$. A calibration based on \cite{bib:calib_imu} was designed and applied to this device. Furthermore, a temperature compensation was designed and implemented.
\item \textbf{Accelerometer:} A 3-axis accelerometer is used to measure gravity. Under the hypothesis that no other accelerations are present, this allows the determination of two of the three Euler angles: $\{\psi, \phi\}$. This hypothesis is acceptable, since the accelerations involved are not significant compared to gravity. A calibration, based on \cite{bib:calib_imu}, was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Magnetometer:} In an area free of magnetic interference this 3-axis sensor will measure $\vec{B}$, the Earth's magnetic field, allowing to determine what direction is North. If the system is horizontal (or the inclination is estimated using other sensors) this sensor can be used to determine the last of the three Euler angles: $\theta$. A calibration based on \cite{bib:bola,bib:alain} was performed on this sensor.
\end{itemize}

The equations used to determine the three Euler angles are the following\cite{bib:kalman}:
  \[
  \phi = -\arcsin\left(\frac{a_x}{\vert\vert\vec{a}\vert\vert}\right), \;  
  \psi =- \arctan \left(\frac{a_y}{a_z}\right)
  \]

  \[
  \hspace{-20pt}
  M = 
\left(\begin{array}{ccc} \cos\!\left(\mathrm{\phi}\right) & \sin\!\left(\mathrm{\phi}\right)\, \sin\!\left(\mathrm{\psi}\right) & \cos\!\left(\mathrm{\psi}\right)\, \sin\!\left(\mathrm{\phi}\right)\\ 0 & \cos\!\left(\mathrm{\psi}\right) & - \sin\!\left(\mathrm{\psi}\right)\\ - \sin\!\left(\mathrm{\phi}\right) & \cos\!\left(\mathrm{\phi}\right)\, \sin\!\left(\mathrm{\psi}\right) & \cos\!\left(\mathrm{\phi}\right)\, \cos\!\left(\mathrm{\psi}\right) \end{array}\right)
  \]

  \[
  \vec{m} = M \; \vec{B}, \;
  \theta = -\arctan\left( \frac{m[1]}{m[0]}\right)
  \]

\subsection{GPS}
\label{sec:sensors-gps}

In theory, given some initial position $\{x_0 ,y_0\}$ the accelerometer could be used to determine variations $\{x_0+\Delta x,y_0+\Delta y\}$. In practice this estimation drifts rapidly (tens of meters in a couple of seconds), so a GPS is used to determine the absolute position $\{x,y\}$ of the system, correcting the drift. The accuracy, with good sky visibility, is of 2-3 meters. The GPS's performance improves when the system is moving.

\subsection{Sensor Specifications}
\label{sec:sensors-specifications}

Table \refp{tab:sensors:resumen} shows an outline of the specifications of the sensors used.

@@

In order to perform adequate control actions, a reliable estimation of the state variables must be available in real time. The Kalman Filter uses the mathematical model for the system to predict what should happen next given the current state, and corrects the prediction with the information read from the sensors, taking into consideration how much confidence is placed on the prediction and how much on the measurements. This weighted prediction-correction technique allows a smooth state estimation without the typical delay introduced by filtering, even small delays can severely affect the performance of the system.

Every sensor has its issues, 
%The gyro drifts over time; the accelerometer is very sensible to the vibrations generated by the motors; the magnetometer is distorted by ferromagnetic materials; the GPS has a very poor accuracy and a slow update rate. Each sensor by itself is very limited,
but they can be combined to compensate for their limitations. The filter takes care of this by integrating all sensors in order to obtain a more accurate state estimation. Figure \refp{fig:diagrama_kalman} shows a diagram of how data from the sensors is combined within the filter (EKF), assisted by the model of the system.
\begin{figure}[h!]
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/diagrama_kalman_eng.png}
	\caption{\textbf{EKF - } Outline of how sensor data is combined to estimate the state variables.}
	\label{fig:diagrama_kalman}
\end{figure}
Euler angles are primary estimated from the combination of Magnetometer and Accelerometer measurements and used as one of the filter inputs. From that estimation of Euler angles and the Accelerometer measurements, the quadrotor velocities $\{v_{qx},\;v_{qy}\;v_{qz}\}$ can be deduced by subtracting gravity and an integration, which are also introduced to the filter. Angular velocities referenced to the quadrotor and height can be easily deduced from Gyroscope and Barometer measurements respectively, which are also introduced to the filter. Furthermore the GPS measurements can be also used as filter input. All this primary estimations from sensors are considered as measurements in the Kalman Filter, while the prediction is based, as said, in the dynamic model. The output of the filter is an estimation of all the variables of the state vector.\\

%\subsection{Extended Kalman Filter}
%\label{sec:kalman-ekf}

>>>>

%\vspace*{-.5cm}
\begin{abstract}
%\boldmath
This paper describes the design and integration of an instrumentation, sensor fusion and control system that allows the autonomous flight of a quadrotor. A comercial frame is used, a mathematical model for the quadrotor is developed and its parameters determined from the characterization of the unit. An intelligence is integrated to play the role of the flight controller. A 9 degrees of freedom Inertial Measurement Unit (IMU) equipped with a barometer is calibrated and added to the platform. Sensor fusion is done by two modified Extended Kalman Filters (EKF): one combining data provided by IMU and the other also including the information provided by GPS. A reliable estimation of the state variables is obtained. Three states representing systematic bias in the accelerometer measurements are also added to the EKF, which improves the inertial estimation of the position. The control actions are obtained from a proportional-integral controller based on an improved Linear Quadratic Regulator (LQR) algorithm. A stable autonomous platform is achieved.%For windless environments, a stable platform is achieved.
% -cambio-
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

There is an important growth in the interest about unmanned aerial vehicles (UAV) in relation to its capabilities to perform a wide spectrum of tasks as monitoring, surveillance, aerial photography, exploration, delivery, rescue, remote sensing, etc.
The continuous advances in low power MEMS sensor devices, low power embedded processors, energy storage through efficient electric batteries and electrical motors technologies have been driving forces in the process of implementation of new and more sophisticated devices of this kind, particularly miniature flying robots. Quadrotors UAV is one of the most popular architectures.
They emerged also as a typical platform for research. There is much research work about path planning and vision towards navigation in unstructured scenarios (e.g.: \cite{bib:tio_cuatro, bib:tio_cinco, bib:tio_seis}), fault tolerant navigation (e.g.:\cite{bib:tio_tres}), path tracking (e.g.: \cite{bib:tio_ocho}), control techniques for UAV such as neural networks \cite{bib:tio_uno}, gain-schedulling \cite{bib:tio_tres}, feedback linearization (e.g.:\cite{bib:tio_seis, bib:tio_siete}), PI control \cite{bib:tio_ocho}, Adaptive Control \cite{bib:tio_dos}, LQR (e.g.: \cite{bib:tio_cinco, bib:lqrnotes}), backstepping control (e.g.:\cite{bib:tio_cuatro}, \cite{bib:tio_seis}), etc.
If state estimation, path planning or control techniques are to be experimented in a physical setup, an important issue is to have the possibility to gain access to the different blocks of the navigational system. 
The purpose of the reported work is to develop the instrumentation and control of an autonomous quadrotor in which is possible to modify and test different state estimation techniques, path planning algorithms and control laws. The solution was reached using a commercial mechanical platform, and design and build an ad-hoc instrumentation and control navigation system.
An aerial vehicle is inherently unstable, staying still is not a simple task. This paper focuses on the stabilization of a quadrotor, explaining the development of the mathematical model to represent the system, the filtering techniques applied for sensor data fusion and the control system that allows the quadrotor to fly.
\begin{figure}
%\vspace{-.8cm}
	\centering
	\includegraphics[width=.8\columnwidth]{./pics_paper/javi.png}
	\caption{\textbf{Commercial motorized-frame used.}}
	\label{fig:commercial_quadrotor}
\end{figure}
\vspace{0.2cm}
\begin{figure}
	\centering
	\includegraphics[width=.8\columnwidth]{./pics_paper/diagrama_general.png}
	\caption{\textbf{General diagram.}}
	\label{fig:diagrama_tio}
	\vspace{-0.5cm}
\end{figure}
\section{System architecture}
The platform is based on the commercial radio controlled quadrotor shown in Figure \refp{fig:commercial_quadrotor}. The length between opposite propellers is $61.5cm$, the weight is $990g$ (including battery), and it has $1300g$ of payload. The frame, motors and the Electronic Speed Controllers (ESCs) used for the motors were preserved, whereas the IMU and intelligence were replaced by the flight controller that was developed. A BeagleBoard\footnote{BeagleBoard development board - \url{http://beagleboard.org/}} running Linux\footnote{Angstrom distribution: \url{http://www.angstrom-distribution.org/}} performs the computations required to convert raw data received from the IMU\footnote{\textit{Mongoose} IMU - \url{http://store.ckdevices.com/}} over a UART and combine it using an EKF. The EKF overcomes the problems inherent to each sensor and filters out noise, providing a reliable estimation of the state vector. Once the current state is known, the LQR algorithm is used to derive the control actions required to bring the system to the desired set-point. A complete diagram of the implemented system is shown in Figure \refp{fig:diagrama_tio}.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill January 11, 2007
The two main goals are to integrate additional sensors and intelligence to the available platform to obtain a state estimation, and design and integrate a control system that, using the state estimation, achieves the autonomous flight.
\section{Model of a quadrotor}
\label{sec:modelo}
\begin{figure}
	\centering
	\includegraphics[width=.7\columnwidth]{./pics_paper/quad_coord.jpg}
	\vspace{-5pt}
	\caption{\textbf{Model of the quadrotor -} The blue arrows represent the inertial reference frame $S_I$, and the red arrows represent the non-inertial reference frame $S_q$. The cyan ``looped'' arrows indicate the direction of rotation of each motor, which rotate at $\omega_i$ and generate a torque $M_i$ opposite to their direction of rotation. The arrows labeled $T_{[1,2,3,4]}$ represent the thrust of the motors. The semicircle and the two yellow spheres indicate the $x_q$ axis of the unit.}
    %\vspace{-10pt}
	\label{fig:quad}
\end{figure}

\subsection{Definitions}
\label{sec:modelo-defs}
A diagram of the quadrotor is shown in Figure \refp{fig:quad}. Two of the motors rotate clockwise (2 and 4) and the other two (1 and 3) rotate counterclockwise. This configuration allows the quadrotor to rotate, tilt and gain/lose altitud by setting different speeds on each motor. Two frames of reference (Figure \refp{fig:quad}) are constantly used through out this paper: an intertial frame $S_I - \lbrace \hat{i},\hat{j},\hat{k}\rbrace$ ($\lbrace \vec{x},\vec{y},\vec{z}\rbrace$), relative to the Earth, mapped to  North, West and Up respectively, and a non-inertial frame $S_q - \lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$ ($\lbrace \vec{x}_q,\vec{y}_q,\vec{z}_q\rbrace$) relative to the quadrotor. The mapping of one frame to the other can be achieved by applying the three rotations shown in Figure \refp{fig:rotaciones}. The angles $\lbrace \theta, \varphi, \psi\rbrace$ are known as Euler angles.

\subsection{Dynamics-kinematics of the system}
\label{sec:modelo-dyn-kin}
From a detailed analysis of the dynamics and kinematics of the quadrotor, the equations \refp{eq:modelo} are obtained, and the state vector shown in \refp{eq:sv} is built to describe the system at any given time. The variables with subscript \textit{q} are referenced to the quadrotor frame $S_q$, the rest are relative to $S_I$:
\begin{equation}
\small
\boxed{\begin{aligned}&\dot{x}\!=\!v_{q_x} \cos \varphi \cos \theta \!+\! v_{q_y} ( \cos \theta \sin \varphi \sin \psi\!-\!\cos \varphi \sin \theta )\\
&\quad \!+\! v_{q_z}(\sin \psi \sin \theta \!+\! \cos \psi \cos \theta \sin \varphi)\\
&\dot{y}\!=\!v_{q_x} \cos \varphi \sin \theta \!+\! v_{q_y} (\cos \psi \cos \theta \!+\! \sin \theta \sin \varphi \sin \psi)\\
&\quad\!+\! v_{q_z}( \cos \psi \sin \theta \sin \varphi\!-\!\cos \theta \sin \psi )\\
&\dot{z}\!=\! \!-\!v_{q_x} \sin \varphi  \!+\! v_{q_y} \cos \varphi \sin \psi  \!+\! v_{q_z}\cos \varphi \cos \psi\\
&\dot{\psi}\!=\!\omega_{q_x} \!+\! \omega_{q_z}\tan\varphi \cos\psi \!+\! \omega_{q_y}\tan\varphi \sin\psi\\
&\dot{\varphi}\!=\!\omega_{q_y}\cos \psi \!-\! \omega_{q_z}\sin\psi\\
&\dot{\theta}\!=\!\omega_{q_z} \frac{\cos\psi}{\cos\varphi}  \!+\! \omega_{q_y}\frac{\sin\psi}{\cos\varphi}\\
&\dot{v_{q_x}}\!=\!v_{q_y} \omega_{q_z} \!-\! v_{q_z} \omega_{q_y}\!+\!g\sin\varphi\\
&\dot{v_{q_y}}\!=\!v_{q_z} \omega_{q_x} \!-\! v_{q_x} \omega_{q_z}\!-\!g\cos\varphi\sin\psi\\
&\dot{v_{q_z}}\!=\!v_{q_x} \omega_{q_y} \!-\! v_{q_y} \omega_{q_x}\!-\!g\cos\varphi\cos\psi\!+\!\frac{1}{M}\sum_{i\!=\!1}^4T_i\\
&\dot{\omega_{q_x}}\!=\!\frac{1}{I_{xx}}\omega_{q_y}\omega_{q_z}(I_{yy}\!-\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{xx}}\omega_{q_y}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{xx}}dMg\cos\varphi\sin\psi\!+\!\frac{1}{I_{xx}}L(T_2\!-\!T_4)\\
&\dot{\omega_{q_y}}\!=\!\frac{1}{I_{yy}}\omega_{q_x}\omega_{q_z}(\!-\!I_{xx}\!+\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{yy}}\omega_{q_x}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{yy}}dMg\sin\varphi\!+\!\frac{1}{I_{yy}}L(T_3\!-\!T_1)\\
&\dot{\omega_{q_z}}\!=\!\frac{1}{I_{zz}}\left(\!-\!Q_1\!+\!Q_2\!-\!Q_3\!+\!Q_4\right)\\
%\frac{\!-\!I_{zz_m}(\dot{w_1}\!-\!\dot{w_2}\+!\!\dot{w_3}\!-\!\dot{w_4})}{I_{zz}}\!+\!
\end{aligned}}
\label{eq:modelo}
\end{equation}
\begin{equation}
\mathbf{x}=\left\lbrace  x,y,z, \theta,\varphi,\psi, v_{q_z},v_{q_y},v_{q_z},\omega_{q_x},\omega_{q_y},\omega_{q_z} \right\rbrace
\label{eq:sv}
\end{equation}
where:
\begin{itemize}
\item $\lbrace x,y,z \rbrace$ represent the position of the center of mass of the system in $S_I$.
\item $\lbrace\theta,\varphi,\psi\rbrace$ are the Euler angles shown in Figure \refp{fig:rotaciones}.
\item $\lbrace v_{q_x},v_{q_y},v_{q_z}\rbrace$ are the linear velocities relative to $S_q$.
\item $\lbrace w_{q_x},w_{q_y},w_{q_z}\rbrace$ are the angular velocities relative to $S_q$ (right hand rule applied on $\lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$).
\item $T_1(\omega_1),\;T_2(\omega_2),\;T_3(\omega_3),\;T_4(\omega_4)$ are the thrust of the motors.
\end{itemize}
The dynamical model considered, derived from equations \ref{eq:modelo}, is
\begin{equation}
	\dot{\mathbf{x}}=\mathcal{F}(\mathbf{x},\mathbf{u})
	\label{ec:fxu}
\end{equation}
where $\mathbf{u}=\{\omega_1,\omega_2,\omega_3,\omega_4\}$ is the controllable input of the system. 
\begin{figure}[h!]
  \centering
  \vspace{-10pt}
  \subfloat[Rotation 1: Axis $\hat{k}$]{\label{fig:angulos_1}
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_1.png}}
  \subfloat[Rotation 2: Axis $\hat{j}$]{\label{fig:angulos_2} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_2.png}}
  \subfloat[Rotation 3: Axis $\hat{i}$]{\label{fig:angulos_3} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_3.png}}
  \caption{\textbf{Mapping -} Rotations applied on $S_I$ to obtain $S_q$.}
  \label{fig:rotaciones}
\end{figure}

The mathematical model developed is similar to the ones presented in \cite{bib:quadrotor-bible, bib:curso_quad}, but also takes into account that the center of gravity of the quadrotor is not at the same height as the propellers. Thus a momentum produced by the gravity force has to be added. While in \cite{bib:curso_quad} the linear velocities are expressed in an inertial frame, in this work are referenced to the quadrotor frame. This choice simplifies the theoretical development and the interpretation of the data provided by the IMU, which is mounted on the quadrotor and hence provides accelerations and angular velocities that are relative to $S_q$.

@@

The IMU is equiped with the following sensors:
\begin{itemize}
\item \textbf{Barometer:} Measures the absolute pressure of the environment. Variations of pressure are used to estimate variations in the altitude of the system.
\item \textbf{Thermometer:} The barometer includes a thermometer. The temperature data is used to apply a temperature compensation to the calibrations performed on the gyroscope and the accelerometer. The reader can find the details in this procedure in \cite{bib:nosotros}.
\item \textbf{Gyroscope:} A 3-axis gyroscope is used to measure angular velocity of $S_q$. A calibration based on \cite{bib:calib_imu} was designed and applied to this device. Furthermore, a temperature compensation was designed and implemented.
\item \textbf{Accelerometer:} A 3-axis accelerometer is used to measure gravity. Under the hypothesis that no other accelerations are present, this allows the determination of two of the three Euler angles: $\{\psi, \phi\}$. This hypothesis is acceptable, since the accelerations involved are not significant compared to gravity. A calibration, based on \cite{bib:calib_imu}, was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Magnetometer:} In an area free of magnetic interference this 3-axis sensor will measure $\mathbf{B}$, the Earth's magnetic field, allowing to determine what direction is North. If the system is horizontal (or the inclination is estimated using other sensors) this sensor can be used to determine the last of the three Euler angles: $\theta$. A calibration based on \cite{bib:bola,bib:alain} was performed on this sensor.
\end{itemize}

\subsection{GPS}
\label{sec:sensors-gps}
In theory, given some initial position $\{x_0 ,y_0\}$ the accelerometer could be used to determine variations $\{x_0+\Delta x,y_0+\Delta y\}$. In practice this estimation drifts rapidly (ten meters in about ten seconds), so a GPS is used to determine the absolute position $\{x,y\}$ of the system, correcting the drift. The accuracy, with good sky visibility, is of 2-3 meters. The GPS's performance improves when the system is moving.

\subsection{Sensor Specifications}
\label{sec:sensors-specifications}
Table \refp{tab:sensors:resumen} shows an outline of the specifications of the sensors used.

@@

In order to perform adequate control actions, a reliable estimation of the state variables must be available in real time. The Kalman Filter uses the mathematical model for the system to predict what should happen next given the current state. It corrects the prediction with the information read from the sensors, taking into consideration how much confidence is placed on the prediction and how much on the measurements. This weighted prediction-correction technique allows a smooth state estimation without the typical delay introduced by filtering, even small delays can severely affect the performance of the system.

Every sensor has its issues, the accelerometer drifts over time; the gyroscope is very sensible to the vibrations generated by the motors; the magnetometer measure is distorted by ferromagnetic materials; the GPS has a very poor accuracy and a slow update rate. Each sensor by itself is very limited, but they can be combined to compensate for their limitations. The filter takes care of this by integrating all sensors in order to obtain a more accurate state estimation.

The theory behind a standard Kalman Filter does not hold for a nonlinear system. The model for the quadrotor given by \refp{eq:modelo} is highly nonlinear, so an Extended Kalman Filter (EKF) is implemented. Several authors (e.g. \cite{bib:kalman,bib:kalman2}) used EKF to overcome this difficulty. In this work a modified EKF was developed, similar to \cite{bib:kalman,bib:kalman2}, but with 3 states added representing the accelerometer bias, which improved substantially the linear velocities and position estimation. While the Kalman Filter ensures a statistical optimal performance, the EKF is not optimal, and it is not possible to determine the error a priori, due to high dependency of the performance with the linearization \cite{bib:kay}. Although, EKF is the most used and popular filtering technique in navigation problems.

\subsection{Mathematical model}
